---
title: "Technical report for Crawford school fiscal policy essay"
author: "Thomas McMahon"
date: "29/04/2016"
output:
  html_document: default
  pdf_document:
    keep_tex: yes
params:
  cap_level: 10000
  r: .045
  loss: .8
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ozTaxData)
library(knitr)

# income tax and extra revenue functions
source("income_tax.R")

# Data setup
tax <- sample_13_14
tax$WRE_total <- tax$WRE_car_amt + tax$WRE_trvl_amt + tax$WRE_uniform_amt +
                 tax$WRE_self_amt + tax$WRE_other_amt
tax$deciles <- ntile(tax$Tot_inc_amt, 10)
cap_level <- params$cap_level
tax$cap_level <- cap_level
r <- as.numeric(params$r)
loss <- as.numeric(params$loss)

# Helper Functions
millions <- function(x) paste("$",round(x / 1e6, 0), "m", sep = "")
pop_total <- function(x, t = 1) (x * 50) * (1 + r) ^ t

```

## Data

The data source for this modelling is the ATO's two per cent sample of tax
returns for the 2013-14 financial year. The ATO provide researchers a
representative sample of individual taxpayers' return data, with some
modifications to protect individuals' privacy. The sample file contains
demographic information, as well as information on income and deductions claimed
for work related expenses.

The sample file only contains information about taxpayers that lodged a tax
return, and therefore is not representative of the population as a whole.
Therefore the population for analysis will be taxpayers, rather than all
Australians. For the purposes of analysing the effects of a cap to work related
expenses this is a relatively minor problem, as only people that lodge tax
returns claim these deductions.

As the most recent data is from 2013-14, I will adjust the data for wages growth
[XX per cent] and population growth [XX per cent].

## Methodology

This analysis will follow a static or 'day-after' static micro-simulation 
modelling approach, this does not take account of any behavioural changes or 
second-round effects that will result from the introduction of the cap. Static 
micro-simulation is a technique commonly used to analyse changes in tax or 
transfer policy. In Australia there are a number of these models, including the 
National Centre for Social and Economic Modelling's STINMOD model, in Europe the
EUROMOD model is an example of tax-benefit micro-simulation model.

As I don't have access to the STINMOD model, I will develop a simple 
micro-simulation model using the ATO tax data sample. This approach has been 
used in Australia by Phillips to model changes to [tax deduction rules for 
property
investors](http://rsss.anu.edu.au/sites/default/files/Distributional_Modelling%20_Negative_Gearing_and_Capital_Gains.pdf)
and [average tax rates across income 
distributions](http://rsss.anu.edu.au/sites/default/files/Average_Tax_Rate_by_Income_Distribution_Australia.pdf).
I will follow the methodology described by Phillips in this analysis.

To calculate the tax value of the benefit of deductions over the cap, I 
calculated the tax liability for each individual in the tax return data. This 
analysis does not take into account all of the features of the income tax 
system, but it does feature the major aspects including personal income tax, the
medicare levy and the low income tax offset, this mirrors the approach adopted 
by Phillips. Then I analysed the tax owed both with and without capped
deductions. The lost tax benefit is the difference between what the individual
could deduct with and without the cap. This result was then weighted to obtain
population totals.

## Fiscal Results

```{r fiscal}
# R spits an error if this is in a single step
# Uses Tot_inc_amt as an approximation of gross income
x <- mapply(revenue_increase, tax$Tot_inc_amt, tax$WRE_total, tax$cap_level)
tax$extra_revenue <- x

revenue <- tax %>% summarise(revenue = sum(extra_revenue))

#revenue <- revenue[[]]

# Create a dataframe to calculate the compound growth
years <- seq(0,5)
fin_year <- c('13-14', '14-15', '15-16', '16-17', '17-18', '18-19')
yearly_rev <- sapply(years, function(x) {
    millions(pop_total(revenue, t = x) * loss ) # [[]] to get the value
    })

fiscal_table <- knitr::kable(data.frame(years, fin_year, yearly_rev))

fiscal_table
```

## Distributional Results

```{r distributional}
# Income
tax %>% 
    select(deciles, WRE_total, extra_revenue) %>% 
    group_by(deciles) %>% 
    summarise(
        # 2017-18 figures
        total_tax = millions(sum(pop_total(extra_revenue, t = 4) * loss)), 
        average_tax = paste("$", round(mean(extra_revenue)), sep = "")) %>%
    kable


# Gender
tax %>% 
    select(Gender, WRE_total, extra_revenue) %>% 
    group_by(Gender) %>% 
    summarise(
        total_tax = millions(sum(pop_total(extra_revenue, t = 4) * loss)),
        average_tax = paste("$", round(mean(extra_revenue)), sep = "")) %>%
    kable
```
